Developmental plasticity to high light intensity amplifies the advantage of amphistomous leaves

### 2025-07-06 ms/ms.qmd
tidying up fitting and plotting

### Bayesian data analysis
n_model = 5 |> # i don't think I need this anymore
  number_to_word()

@@

@@

### Bayesian phylogenetic mixed effects models in *Stan*
#### MARKER - REVISE MODEL DESCRIPTION
We fit `r n_model` models to test predictions of competing hypotheses about why amphistomy advantage (\aax) might be greater for leaves in sunny, open habitats. This section provides an overview of differences among models (\autoref{tbl-models}). The next sections describe how we fit models in *Stan*, all model parameters and priors, and specific predictions about parameter values for each hypothesis.

```{r, models, echo=FALSE,message=FALSE,warning=FALSE}
#| label: tbl-models
#| tbl-cap: "Summary of differences among competing models of how \\aax{} varies with light intensity, light treatment, and among populations as a function of native \\ppfd. The models are numbered from simpler to more complex. All models include fixed effects of light intensity and light treatment; some models include interactions between. All models include a phylogenetic random effect of population on \\aax; some models include varying effects of light intensity and light treatment among populations. The last column indicates the population-level \\aax{} variable we used as a response to native \\ppfd."

tribble(
    ~"Model", ~"Fixed effects", ~"Phylogenetic random effects", ~"Response to native \\ppfd",
    "1", "light intensity", "varying intercept among populations", "$\\mathrm{AA}_{0, \\text{acc}}$",
    "1", "light treatment", "", "",
    "2", "light intensity", "varying intercept among populations", "$\\mathrm{AA}_{0, \\text{acc}}$",
    "2", "light treatment", "", "",
    "2", "intensity $\\times$ treatment", "", "",
    "3", "light intensity", "varying intercept among populations", "$\\mathrm{AA}_{2000, \\text{acc}}$",
    "3", "light treatment", "varying effect of high light intensity among populations", "",
    "4", "light intensity", "varying intercept among populations", "$\\mathrm{AA}_{\\text{sun, acc}}$",
    "4", "light treatment", "varying effect of sun treatment among populations", "",
    "5", "light intensity", "varying intercept among populations", "$\\mathrm{AA}_{2000, \\text{sun, acc}}$",
    "5", "light treatment", "varying effect of high light intensity among populations", "",
    "5", "intensity $\\times$ treatment", "varying effect of sun treatment among populations", ""
  ) |>
  knitr::kable(
    format = "latex",
    align = c("c", "l", "l", "l"),
    booktabs = TRUE,
    escape = FALSE,
    longtable = FALSE
  ) |>
    kableExtra::collapse_rows(columns = 1, latex_hline = "major", row_group_label_position = "first") |>
  kableExtra::column_spec(1, width = "0.3in") |>
  kableExtra::column_spec(2, width = "1.3in") |>
  kableExtra::column_spec(3, width = "3.4in") |>
  kableExtra::column_spec(4, width = "1in")

#### Fitting models in *Stan* {#sec-fitting}

We fit Bayesian models with MCMC sampling in the probabilistic programming language *Stan* [@stan_development_team_stan_2025] using the *R* package **brms** version `r fit_aa1$version$brms` [@burkner_brms_2017]. We used *CmdStan* version `r fit_aa1$version$cmdstan` and **cmdstanr** version `r fit_aa1$version$cmdstanr` [@gabry_cmdstanr_2025] to interface with *R* version `r getRversion()` [@r_core_team_r:_2025]. We sampled the posterior distribution from `r dim_fit_aa[2]` chains with `r dim_fit_aa[1]` iterations each after `r dim_fit_aa[1]` warmup iterations per chain. We estimated parameters and confidence intervals as the median and 95% quantile intervals of the posterior, respectively. We chose the number of chains, warmup and sampling iterations, and maximum treedepth so that parameter estimates converged ($\hat{R} < `r aa_args['max_rhat']`$ [@gelman_inference_1992]) and the effective sample size (ESS) for each parameter was $> `r aa_args['min_ess']`$.

#### Parameter estimation and priors {#sec-parameters}

There were four levels of parameter estimation in our analysis:

1. Estimate \agcurve{} curve parameters
2. Estimate \aax{} for each light intensity with leaf using \agcurve{} curve parameters
3. Estimate the effects of light intensity, light treatment, and population on \aax{} (assimilatory and plasticity hypotheses)
4. Estimate the effects of native light habitat on population-level \aax{} (constitutive hypothesis)

Although the higher-level parameter estimates depend on the lower-level parameter estimates, we fit all models simultaneously to ensure that the uncertainty in lower-level estimates propagated to higher-level estimates.

```{r, parameters, echo=FALSE,message=FALSE,warning=FALSE}
#| label: tbl-parameters
#| tbl-cap: "Description of parameters estimated in the hierarchical Bayesian model. The `Parameter` column lists the parameter name as it appears in text. The `Description` column provides a brief description of the parameter."

df_parameters = read_csv("../data/parameters1.csv", col_types = "ccccl") |>
    dplyr::filter(include)

df_parameters |>
  dplyr::select(Parameter = latex, Description = description) |>
  knitr::kable(
    format = "latex",
    align = c("l", "l"),
    booktabs = TRUE,
    escape = FALSE,
    longtable = TRUE
  ) |>
  kableExtra::kable_styling(latex_options = c("striped", "no_vertical_lines")) |>
  kableExtra::column_spec(1, width = "1in") |>
  kableExtra::column_spec(2, width = "5in") |>
  kableExtra::pack_rows("\\agcurve{} curve parameters",
                        min(which(df_parameters$par_type == "ags")),
                        max(which(df_parameters$par_type == "ags")),
                        escape = FALSE) |>
  kableExtra::pack_rows("\\aax{} for each light intensity with leaf using \\agcurve{} curve parameters",
                        min(which(df_parameters$par_type == "aa_est")),
                        max(which(df_parameters$par_type == "aa_est")),
                        escape = FALSE) |>
  kableExtra::pack_rows("effects of light intensity, light treatments, and population on \\aax",
                        min(which(df_parameters$par_type == "aa")),
                        max(which(df_parameters$par_type == "aa")),
                        escape = FALSE) |>
  kableExtra::pack_rows("effects of native light habitat on population-level \\aax",
                        min(which(df_parameters$par_type == "ppfd_aa")),
                        max(which(df_parameters$par_type == "ppfd_aa")),
                        escape = FALSE)

##### \agcurve{} curve parameters

We modeled \logA as a quadratic function of \loggsw for each leaf using the following equation:

$$\log(A_{ij}) = (\beta_0 + b_{0,j}) + (\beta_1 + b_{1,j}) \log{g_{\text{sw},i}} + (\beta_2 + b_{2,j}) \log{g_{\text{sw},i}}^2 + \epsilon_{i}$$

where $\beta_0$, $\beta_1$, and $\beta_2$ are the average intercept, linear, and quadratic coefficients, respectively. We used diffuse normal priors with mean $0$ and standard deviation $10$ on these parameters. We estimated random effects of curve $j$ on the intercept ($b_{0,j}$), linear ($b_{1,j}$), and quadratic ($b_{2,j}$) coefficients. We assumed that the $j \times 3$ array of coefficients was multivariate normal a mean vector of $\vec{0}$ and covariance $\Sigma_\text{curve}$. We used a weakly informative normal prior with mean $0$ and standard deviation $1$ on the log-transformed standard deviations (i.e. the diagonal of $\Sigma_\text{curve}$). We used a weakly informative $\mathrm{LJK}(2)$ prior on the correlation matrix. The off diagonal elements of $\Sigma_\text{curve}$ can be calculated from its diagonal elements and the correlation matrix.

The residuals $\epsilon_{i}$ were modeled as a lag-1 autocorrelated time-series. We further assumed that the residual standard deviation of the $j^{\text{th}}$ curve ($\sigma_{\epsilon,j}$) was inversely proportional to the leaf surface area ($S_j$) within the chamber:

$$\log (\sigma_{\epsilon,j}) = \log (\sigma_{\qty{6}{\raiseto{2}\centi\meter},\epsilon}) + (6 - S_j) \beta_{S,\epsilon}$$
where $\sigma_{\qty{6}{\raiseto{2}\centi\meter},\epsilon}$ is the minimum residual standard deviation when the $\qty{6}{\raiseto{2}\centi\meter}$ chamber is completely filled. The residual standard deviation increases on log-linear scale by $\beta_{S,\epsilon}$. We used a weakly informative normal prior with mean $-3$ and standard deviation $5$ on $\log (\sigma_{\qty{6}{\raiseto{2}\centi\meter},\epsilon})$ and a weakly informative normal prior with mean $0$ and standard deviation $1$ on $\beta_{S,\epsilon}$.

##### \aax{} for each light intensity with leaf using \agcurve{} curve parameters

Within the $k^{\text{th}}$ leaf, we estimated \aax{} for each light intensity by integrating the difference in \logA{} between the amphi and pseudohypo \agcurve{} curves over the range of \gsw{} values where the curves overlap (from $\min(\log(g_\text{sw}))$ to $\max(\log(g_\text{sw}))$). The estimate of \aax{} for the $k^{\text{th}}$ leaf at light intensity $l$ in population $m$ is:

$$\widehat{\mathrm{AA}}_{klm} = \int_{\text{min(log(}g_\text{sw}))}^{\text{max(log(}g_\text{sw}))} \text{log}\bigg(\frac{\hat{A}_\text{amphi}(x; \theta_{klm,\text{amphi})}}{\hat{A}_\text{hypo}(x; \theta_{klm,\text{hypo})}}\bigg) dx$$
where:

$$\theta_\text{amphi} \in \{\hat{b}_{0, f(\text{amphi}, k,l,m)}, \hat{b}_{1, f(\text{amphi}, k,l,m)}, \hat{b}_{2, f(\text{amphi}, k,l,m)}\}, \text{and}$$

$$\theta_\text{hypo} \in \{\hat{b}_{0, f(\text{hypo}, k,l,m)}, \hat{b}_{1, f(\text{hypo}, k,l,m)}, \hat{b}_{2, f(\text{hypo}, k,l,m)}\}.$$ 
The function $f : \Theta_1 \rightarrow \Theta_2$ maps the set $\Theta_1$ indexed by leaf type (amphi or pseudohypo), leaf replicate, light intensity, and population to set $\Theta_2$ indexed by individual \agcurve{} curve. This mapping is necessary because the random effects structure differs between models of \loggsw{} on \logA{} and that of models predicting \aax{} described in the next section.

@@

* Variation in stomatal ratio among pops and light treatments
* Table S7: Model comparison LOOIC tables
* Table S8: Model comparison LOOIC tables with LMA
* Table S9: Calculation of alternate scenario where amphi leaves can get same A with less R

>>>>

### Bayesian data analysis
n_model1 = nrow(aa_loo_table1)
n_model2 = nrow(aa_loo_table2)

@@

note: estimates of AA are less than that described by Parkhurst and Mott (1990) and also less than that of Xiong and Flexas (2021). What does this tell us?

@@

### Bayesian data analysis with *Stan* {#sec-analysis}

Our analysis pipeline consisted of four main steps:

1. Estimate \agcurve{} curve parameters for each leaf
2. Estimate \aax{} for both measurement light intensities using \agcurve{} curve parameters
3. Estimate the effects of light intensity, light treatment, and population on \aax{} (assimilatory and plasticity hypotheses)
4. Estimate the effects of native light habitat on population-level \aax{} (constitutive hypothesis)

All models were using a Bayesian model with HMC sampling in the probabilistic programming language *Stan* [@stan_development_team_stan_2025] using the *R* package **brms** version `r fit_aa1$version$brms` [@burkner_brms_2017]. We used *CmdStan* version `r fit_aa1$version$cmdstan` and **cmdstanr** version `r fit_aa1$version$cmdstanr` [@gabry_cmdstanr_2025] to interface with *R* version `r getRversion()` [@r_core_team_r:_2025]. We sampled the posterior distribution from a single chain with a minimum of `r aa_args$n_iter_init / 2` iterations after `r aa_args$n_iter_init / 2` warmup iterations. If necessary, we refit models with more iterations to decrease the Gelman-Rubin convergence statistic ($\hat{R}$) [@gelman_inference_1992] to less than $`aa_stats[["max_rhat"]]`$ and the bulk effective sample size (ESS) to greater than $`aa_stats[["min_ess"]]`$ for all model parameters. We also increased the the `adapt_delta` parameter until there were fewer than $`aa_stats[["n_divergent"]]`$ divergent transitions.

#### \agcurve{} curve parameters

We modeled \logA as a quadratic function of \loggsw for each leaf, measured at low and high light intensity, in both amphistomatous (untreated) and pseudohypostomatous (propafilm treatment) states. At minimum, each model was fit using four chains with 2000 warmup and sampling iterations per chain. In cases where we encountered divergent transitions while fitting, we increased the number of sampling iterations and `adapt_delta` parameter until there were fewer than 10 divergences.

#### \aax{} for each light intensity within leaf using \agcurve{} curve parameters
##### MARKER - add figure of this
We estimated \aax{} for each light intensity by integrating the difference in \logA{} between the amphi and pseudohypo \agcurve{} curves over the range of \gsw{} values where the curves overlap (from $\min(\log(g_\text{sw}))$ to $\max(\log(g_\text{sw}))$).

$$\widehat{\mathrm{AA}} = \int_{\text{min(log(}g_\text{sw}))}^{\text{max(log(}g_\text{sw}))} \text{log}\bigg(\frac{\hat{A}_\text{amphi}(x; \theta_{\text{amphi}})}{\hat{A}_\text{hypo}(x; \theta_{\text{hypo}})}\bigg) dx$$
where $\theta_\text{amphi}$ and $\theta_\text{hypo}$ are the quadratic parameters of the amphi and pseudohypo \agcurve{} curves, respectively. We calculated $\widehat{\mathrm{AA}}$ from each draw of the posterior for the paired amphi and pseudohypo \agcurve{} curves to obtain the posterior distribution of \aax{} for each leaf at each light intensity. We then summarized the posterior distribution of \aax{} for each leaf at each light intensity by the posterior median, which became our estimate of \aax{} for subsequent analyses.

### Bayesian phylogenetic mixed effects models

We fit Bayesian mixed effects models with phylogenetically structured random effects to test predictions of competing hypotheses about why amphistomy advantage (\aax) might be greater for leaves in sunny, open habitats. We fit all models combinations of measurement light intensity, growth light intensity, and their interaction as both fixed effects and random effects among populations. In other words, we test whether there were main effects of light intensity treatments and/or whether populations varied in their response to light intensity treatments. In all models, we included population as a phylogenetically structured random effect. We crossed those fixed and random effect structures with fixed effects of measurement light intensity and/or growth light intensity on the distributional parameter, $\sigma$, the residual variance in \aax{}. In total, we fit `r n_model1` models, each with a different combination of fixed, random, and distributional effects. We used the RAxML whole-transcriptome concatenated phylogeny based on 2,745 100-kb genomic windows from @pease_phylogenomics_2016. Two of our populations were not in this tree. We used accession LA1044 in place of LA3909, two populations of *S. galapagnse*; LA0750 was added as sister to LA0716, two closely replated populations of *S. pennellii*. The node was placed half-way between the next deepest node.

NEED TO EDIT THIS: The next sections describe how we fit models in *Stan*, all model parameters and priors, and specific predictions about parameter values for each hypothesis.
Each curve was fit For initial fits, we sampled the posterior distribution from a single chain with `r aa_args$n_iter_init / 2` iterations after `r aa_args$n_iter_init / 2` warmup iterations. After model selection (see below), we refit the selected model with more iterations if necessary to decrease the Gelman-Rubin convergence statistic ($\hat{R}$) [@gelman_inference_1992] to less than $`aa_stats[["max_rhat"]]`$ and the bulk effective sample size (ESS) to greater than $`aa_stats[["min_ess"]]`$ for all model parameters.

#### Fitting models in *Stan* using **brms** interface {#sec-fitting}

We fit Bayesian models with HMC sampling in the probabilistic programming language *Stan* [@stan_development_team_stan_2025] using the *R* package **brms** version `r fit_aa1$version$brms` [@burkner_brms_2017]. We used *CmdStan* version `r fit_aa1$version$cmdstan` and **cmdstanr** version `r fit_aa1$version$cmdstanr` [@gabry_cmdstanr_2025] to interface with *R* version `r getRversion()` [@r_core_team_r:_2025]. For initial fits, we sampled the posterior distribution from a single chain with `r aa_args$n_iter_init / 2` iterations after `r aa_args$n_iter_init / 2` warmup iterations. After model selection (see below), we refit the selected model with more iterations if necessary to decrease the Gelman-Rubin convergence statistic ($\hat{R}$) [@gelman_inference_1992] to less than $`aa_stats[["max_rhat"]]`$ and the bulk effective sample size (ESS) to greater than $`aa_stats[["min_ess"]]`$ for all model parameters.

@@

* Variation in stomatal ratio among pops and light treatments
* Table S7: Model comparison LOOIC tables
* Table S8: Model comparison LOOIC tables with LMA
* Table S9: Calculation of alternate scenario where amphi leaves can get same A with less R
* Each A-gs figure with fitted curves, R2, and AA
